---
title: "Teaching Social Work Statistics with R"
author: "Wes Furlong"
date: "7/3/2019"
output:
  word_document: default
  pdf_document: default
  html_document:
    code_folding: hide
    theme: journal
---

```{r setup, include=FALSE, message=FALSE, results='hide', warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

### Call the packages we'll use for the analysis:
tidyverse, mosaic, janitor, data.table, DataExplorer, SmartEDA, and psych.

```{r results='asis', echo=FALSE, include=FALSE,}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)

library(tidyverse)
library(mosaic)
library(janitor)
library(lattice)
library(data.table)
library(DataExplorer)
library(SmartEDA)
library(psych)

```

### Load the dataset:

```{r}
schools <- read.csv("asheville_school_data.csv", stringsAsFactors = FALSE) 
schools <- schools %>%
  clean_names()
schools <- schools %>% rename(name = i_name)
```

### Let's subset a few variables and take a look at the data:

```{r}
schools_1 <- schools %>%
  select(name, type, stud_teach_ratio, free_dis_lunch, enrollment, district, title_1, avg_score_2018, rank_change)

str(schools_1)
describe(schools_1)
```

### Examine the categorical variables:
```{r}
ExpCatViz(schools_1)
```

### Examine the distribution of the numeric variables:
```{r}
ExpNumViz(schools_1)
```

### Examine the missing values:
```{r}
plot_missing(schools_1)
```


### Let's look at school performance by averages scores in 2018: 

```{r}
mean(avg_score_2018 ~ name, data = schools_1, na.rm= TRUE)
```

### How does title I status affect enrollment figures?  Let's look at the boxplot of enrollment by title I status. 

```{r}
bwplot(enrollment ~ title_1, data = schools_1, na.rm = TRUE)
```


### How do the different types of schools compare on state ranking? 
 Let's look at rank change between 2017 and 2018 for public schools, public charter schools, and public magnet schools. We'll first filter type:

```{r}
schools_1_type <- schools_1 %>%
  filter(type %in% c("Public", "Public, Magnet", "Public, Charter")) 

favstats(rank_change ~ type, data = schools_1_type, na.rm = TRUE)
```


### To introduce multivariable thinking, we could use color or size to map a third (categorical) variable. For example, "How does enrollment affect the student teacher ratio by district? 

```{r}
gf_point(stud_teach_ratio ~ enrollment, data = schools_1, color = ~ district) %>%
  gf_lm()
```

### What are the five highest performing high schools by average scores in 2018?

```{r}
schools_1 %>%
  select(name, avg_score_2018) %>%
  arrange(desc(avg_score_2018))
```

### Answer:
 Evergreen Community Charter, Pisgah Elementary, Hominy Valley Elementary, School of Inquiry and Life Sciences, and Glen Arden Elementary.

### Let's take a look at enrollment size and student-teacher ratio for these schools:

```{r}
top5 <- schools_1 %>%
  select(name, avg_score_2018, enrollment, stud_teach_ratio) %>%
  filter(name %in% c("Evergreen Community Charter", "Pisgah Elementary", "Hominy Valley Elementary", "School of Inquiry and Life Sciences", "Glen Arden Elementary"))
data.table(top5)

```

### How do these same variables compare with the lowest five performing schools?

```{r}
schools_1 %>%
  select(name, avg_score_2018) %>%
  arrange(avg_score_2018)
```

```{r}
bottom5 <- schools_1 %>%
  select(name, avg_score_2018, enrollment, stud_teach_ratio) %>%
  filter(name %in% c("Johnston Elementary", "Community High School", "Oakley Elementary", "Woodfin Elementary", "Erwin Middle"))
data.table(bottom5)
```

### Here is another way to filter the top and bottom values. 
 We will create a dataframe for the top and bottom 5 based on values rather than names and then bind the rows together. 

```{r}
top5_scores <- schools_1 %>%
  select(name, avg_score_2018, enrollment, stud_teach_ratio) %>%
  top_n(5, avg_score_2018)

bottom5_scores <- schools_1 %>%
  select(name, avg_score_2018, enrollment, stud_teach_ratio) %>%
  top_n(-5, avg_score_2018)

top_bottom_5 <- bind_rows(top5_scores, bottom5_scores)

top_bottom_5 <- top_bottom_5 %>%
  arrange(desc(avg_score_2018))

data.table(top_bottom_5)


```

### Does it look like enrollment or student teacher ratio affects school performance?
 
Let's visualize the data. First, we'll look at enrollment and average scores. 

```{r}
ggplot(data = top_bottom_5, aes(enrollment, avg_score_2018, label= name)) +
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  geom_text(angle=45, size= 3, hjust= .4, vjust= -.6) +
  theme_bw()
```

### What about student teacher ratio?

```{r}
ggplot(data = top_bottom_5, aes(stud_teach_ratio, avg_score_2018, label= name)) +
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  geom_text(angle=45, size= 3, hjust= .4, vjust= -.6) +
  theme_bw()
```


### What other variables do you think might affect school performance?  
Let's take a look at the percentage of students on free or reduced lunch. 


```{r}
top5_scores <- schools_1 %>%
  select(name, avg_score_2018, enrollment, stud_teach_ratio, free_dis_lunch) %>%
  top_n(5, avg_score_2018)


bottom5_scores <- schools_1 %>%
  select(name, avg_score_2018, enrollment, stud_teach_ratio, free_dis_lunch) %>%
  top_n(-5, avg_score_2018)

top_bottom_5 <- bind_rows(top5_scores, bottom5_scores)

top_bottom_5 <- top_bottom_5 %>%
  arrange(desc(avg_score_2018))

ggplot(data = top_bottom_5, aes(free_dis_lunch, avg_score_2018, label= name)) +
  geom_point() +
  geom_smooth(method = "lm", se=FALSE) +
  geom_text(angle=45, size= 3, hjust= .4, vjust= -.6) +
  theme_bw()
```












